---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/Reckziegel/tidyvars/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Reckziegel/tidyvars/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Reckziegel/tidyvars/graph/badge.svg)](https://app.codecov.io/gh/Reckziegel/tidyvars)
[![Codecov test coverage](https://codecov.io/gh/Reckziegel/tidyvars/graph/badge.svg)](https://app.codecov.io/gh/Reckziegel/tidyvars)
<!-- badges: end -->


# tidyvars

> A tidy interface for VAR models fitted with `{vars}`.

`tidyvars` makes results from Vector Autoregressive (VAR) models easier
to inspect, manipulate and visualize.

The package does not replace `{vars}` or reimplement VAR econometrics.
`{vars}` performs the statistical calculations; `tidyvars` organizes
the results into consistent tidy objects that work naturally with
`{dplyr}`, `{tidyr}`, and `{ggplot2}`.


## Installation

`tidyvars` is currently under development. The `tv_2.0` version can be
installed from GitHub with `{pak}`:

```r
devtools::install_github("Reckziegel/tidyvars")
```

## Quick start

Model estimation stays exactly where it belongs: in `{vars}`.

```{r setup}
library(tidyvars)
library(ggplot2)

model <- vars::VAR(
  EuStockMarkets,
  p = 2,
  type = "const"
)
```

From this point on, `tidyvars` provides a consistent interface for
working with the fitted model.

## Inspect model results

A fitted `{vars}` model contains several equation-specific and nested
objects. `tidyvars` turns the most common results into tidy tables.

```{r inspection}
tv_tidy(model)

tv_glance(model)

tv_augment(model)
```

`tv_tidy()` returns coefficient estimates by equation,
`tv_glance()` provides equation-level summaries, and `tv_augment()`
aligns observed values, fitted values, and residuals.

The outputs use predictable lower-case `snake_case` column names while
preserving the original variable names from the model.

Because the results are tidy, they can immediately enter a normal
tidyverse workflow:

```{r tidy-workflow}
model |>
  tv_tidy() |>
  dplyr::filter(p_value < 0.05) |>
  dplyr::select(equation, term, estimate, p_value)
```

There is no need to navigate `model$varresult`, combine equation
outputs manually, or reshape them before analysis.

## Dynamic analysis

Impulse response functions and forecast error variance decompositions
follow the same pattern.

```{r dynamic-analysis}
set.seed(123)

irf <- tv_irf(model)
fevd <- tv_fevd(model)
```

An impulse response can be inspected directly as data:

```{r irf-filter}
irf |>
  dplyr::filter(
    impulse == "DAX",
    response == "SMI"
  )
```

Each row identifies the horizon, impulse, response, estimate, and
confidence interval.

The same object can be plotted directly:

```{r irf-plot, fig.width = 9, fig.height = 6}
autoplot(irf, layout = "wrap") +
  theme_tidyvars() +
  palette_tidyvars()
```

FEVD results are also returned in tidy form, with one contribution for
each horizon, response, and shock:

```{r fevd-plot, fig.width = 9, fig.height = 5}
autoplot(fevd) +
  theme_tidyvars() +
  palette_tidyvars()
```

Conceptually:

```text
IRF  → response to a shock through time
FEVD → contribution of each shock to forecast uncertainty
```

Both objects can be filtered, transformed, joined, or plotted using
standard tidyverse tools.

## Forecasting

`tv_predict()` converts forecasts from `{vars}` into a single tidy
object containing historical observations, point forecasts, and
confidence intervals.

Multiple confidence levels can be requested at once:

```{r forecasting}
prediction <- tv_predict(
  model,
  n_ahead = 12,
  level = c(0.50, 0.75, 0.90)
)

prediction
```

The result can be plotted directly:

```{r forecast-plot, fig.width = 9, fig.height = 6}
autoplot(prediction, n_history = 30) +
  theme_tidyvars() +
  palette_tidyvars()
```

`n_history` controls only how much historical context is displayed.
The forecast horizon itself is never truncated.

As with the other `tidyvars` objects, the forecast remains available as
ordinary tidy data for further manipulation.

## Diagnostics

Common VAR diagnostics are exposed through the same tabular interface:

```{r diagnostics}
tv_causality(model)

tv_normality_test(model)

tv_serial_test(model)
```

These functions organize the corresponding results from `{vars}` rather
than reimplementing the underlying statistical tests.

The returned tables use normalized identifiers and predictable schemas,
making diagnostic results easy to filter, compare or include in
reports.

## Visualization

`tidyvars` provides `autoplot()` methods for IRFs, FEVDs, and forecasts.

The plotting methods define the structure of the visualization but do
not impose the visual identity of the package.

A `tidyvars` plot is still a regular `ggplot` object:

```r
autoplot(irf) +
  ggplot2::theme_classic()
```

The optional visual system separates layout from colour:

```r
autoplot(irf) +
  theme_tidyvars() +
  palette_tidyvars()
```

```text
theme_tidyvars()   → layout and typography
palette_tidyvars() → colours
```

`theme_tidyvars()` controls structural elements such as typography,
spacing, grids, facets, and legends without changing the colours used
for the data.

`palette_tidyvars()` applies the package colour identity only when
explicitly requested.

Discrete scales are also available independently for ordinary ggplot2
graphics:

```r
scale_color_tidyvars()
scale_fill_tidyvars()
```

This keeps the visual system optional and fully compatible with normal
ggplot2 customization.

