---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyvars

> "tidiers" for VAR models

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/Reckziegel/tidyvars/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Reckziegel/tidyvars/actions/workflows/R-CMD-check.yaml)
[![CRAN
status](https://www.r-pkg.org/badges/version/tidyvars)](https://CRAN.R-project.org/package=tidyvars)
[![Codecov test
coverage](https://codecov.io/gh/Reckziegel/tidyvars/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Reckziegel/tidyvars?branch=main)

<!-- badges: end -->

`tidyvars` is designed to integrate the `vars` package with the
tidyverse ecosystem: it introduces "tidiers" to simplify the analysis of
Vector Autoregressive (VAR) Models, following the "tidy data"
philosophy. The package makes VAR modelling more accessible and
intuitive, especially for old users of `broom`.

## Why Use tidyvars?

Choosing `tidyvars` for working with VAR Models offers key advantages:

**1. Efficient Data Handling**: `tidyvars` converts VAR outputs into a
"tidy format" simplifying data manipulation.

**2. Tidyverse Integration**: `tidyvars` ensures smooth interoperability
with famous tools, like `dplyr` and `ggplot2`.

**3. Enhanced Interpretability**: `tidyvars` significantly eases the
understanding and interpretation of model results.

<!-- __4. Focus on Analitical Skils__: for tidyverse users, `tidyvars` shifts the focus  -->

<!-- from data wrangling to model analysis.  -->

## Installation

You can install the development version of `tidyvars` from github with:

``` r
# install.packages(devtools)
devtools::install_github("https://github.com/Reckziegel/tidyvars")
```

## Usage Examples

Effortlessly manipulate, summarize, and augment VAR models using
`tv_tidy()`, `tv_glance()`, and `tv_augment()`:

```{r example}
library(tidyvars)
library(ggplot2)

mod <- vars::VAR(EuStockMarkets)

# Tidies the VAR model for easier analysis
tv_tidy(mod)

# Provides a concise summary of the model
tv_glance(mod)

# Augments the original data with model's details
tv_augment(mod)
```

The `tidyvars` functions begins with `tv_*()` and maintain the same
suffix as found in the `vars` package for consistency.

To evaluate the model's adherence to standard parametric assumptions use
`tv_normality_test()` and `tv_serial_test()`:

```{r}
# Tests for multivariate normality of residuals
tv_normality_test(mod)

# Checks for serial correlation in residuals
tv_serial_test(mod)
```

To delve deeper into the causal relationships, see `tv_causality()`:

```{r}
# Analyzes causal relationships in the VAR model
tv_causality(mod)
```

VAR models are specially useful for forecasting given the ability to capture
interdependencies of multiple time series. In `tidyvars`, this task 
be easily accomplished with `tv_predict()`:

```{r}
# Generates forecasts from the VAR model
tv_predict(mod, n.ahead = 1)
tv_predict(mod, n.ahead = 2)
```

For more nuanced inferences about the relationships in
the model, use `tv_fevd()` and `tv_irf()`:

```{r}
# Conducts Forecast Error Variance Decomposition
tv_fevd(mod)

# Computes Impulse Response Functions
tv_irf(mod)

# IRF from Blanchard-Quah Decomposition
tv_irf(vars::BQ(mod))
```

The functions `tv_fevd()` and `tv_irf()` are particularly useful for
understanding how shocks to one variable might propagate through the
system and impact other variables over time.

## Plotting

`tidyvars` integrates the `autoplot()` method for most functions
to enhance the user experience with a quick and intuitive visual
representation of model outputs.

See some examples bellow:

```{r, warning=FALSE}
# Fitted Model
autoplot(tv_augment(mod))
# Visualizes Impulse Response Functions
autoplot(tv_irf(mod))
# Displays Forecast Error Variance Decomposition
autoplot(tv_fevd(mod))
# Generates visual forecasts
autoplot(tv_predict(mod, n.ahead = 12))
# Zoom in the forecasts (last 3 months)
autoplot(tv_predict(mod, n.ahead = 12), .n_obs = 60)
```
